A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-29 paper.tex
minor
Definitions

An abstraction of the game restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Figure~\ref{fig:agt} shows an
example abstract game tree for the controller (abstract game trees for the
environment are similar).  In the abstract game, the controller is required to
pick actions from the tree, starting from the root node.  After reaching a
leaf, it continues playing unrestricted.  The tree in Figure~\ref{fig:agt}
restricts the initial controller action to the set $\{a, d\}$.  After choosing
action $a$, the controller is required to play action $b$ in the next round.
It then reaches a leaf of the tree and continues playing unrestricted.  The
root of the tree is annotated by the initial state $s$ of the abstract game and
the bound $k$ on the number of rounds.  We denote $\textsc{nodes}(T)$ the set
of all nodes of a tree $T$ and $\textsc{leaves}(T)$ the subset of leaf nodes.

@@

The tree is first extended to the maximum bound with edges that are labeled
with arbitrary opponent actions (Algorithm~\ref{alg:bounded}, line 14) as shown
in Figure~\ref{fig:extended}. Then copies of state and action variables are
introduced for each node in tree and opponent action variables for each edge.
Opponent actions are set to $\textsc{action}(e)$, the action that labels the
edge of the tree $e$. The state and action variables of node $n$ are connected
to successor nodes $\textsc{succ}(n)$ by an encoding of the transition relation
and constrained to the winning condition of the player.

@@

\begin{algorithm}
    \caption{Tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{ $\lnot \Call{E}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{E}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{\Call{E}{$X_{T}$}}
        \Else
        \State \Return{ $\Call{E}{X_{T}} \lor$ \\
        $$\bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Require $s(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $s(X_T) \land B^M$
        \Function{learn}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$s, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $s(X_T) \land B^m[$\Call{bound}{$T$}$] \equiv \bot$
        \Function{\textoverline{learn}}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min bound
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{bound}{$n$}}
                \State $B^m[i] \gets B^m[i] \setminus \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$s, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

@@

\begin{algorithm}[t]
    \caption{Amended tree formulas for Controller and Environment}
    \label{alg:unboundedTreeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{$B^M$}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{bound}{T} = 0$}
        \State \Return{\Call{E}{$X_T$}}
        \Else
        \State \Return{ \Call{$B^m[\Call{bound}{T}]$}{$X_{T}$} $\land$ \\
            $$\Call{E}{X_T} \lor \bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

Learning of states losing for the environment is similar (\textsc{\textoverline{learn}} in Algorithm~\ref{alg:learn}). 
The main difference is that environment-losing states are only losing for a particular
bound: given that there does not exist an environment strategy that
forces the game into an error state in $k$ rounds or less; there may still exist a longer
environment-winning strategy.  We therefore record learned environment-losing states
along with associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.  $B^m[k]$ 
are initialized to $\top$ for all $k$.  Whenever an environment-losing set $\mathcal{I}$
is discovered for a node $n$ with bound $\textsc{bound}(n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is subtracted from 
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

@@

Formally, $\II$ has the property $\II \land F_2 \equiv \bot$.  Since $T_2$ is
of depth 1, this means that the environment cannot force the game into
$B^m[\textsc{bound}(n)]$ playing against the counterexample moves in $T_2$.
Hence, $\II \cap Upre(B^m[\textsc{bound}(n)]) = \emptyset$.  Furthermore, since
the may-invariant holds, $\II \cap Upre(i) = \emptyset$, for all $i <
\textsc{bound}(n)$.  Hence, removing $\II$ from all $B^m[i], i\leq
\textsc{bound}(n)$ in line~(15) preserves the may-invariant, thus satisfying the
first post-condition.

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ also satisfies $\II$.  Hence,
removing $\II$ from $B^m[\textsc{bound}(n)]$ makes $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.

At the second last recursive call to \textsc{\textoverline{learn}}, tree $T_1$
is empty, $n$ is the root node, $\textsc{\textoverline{treeFormula}}(T_1)
\equiv B^m[\textsc{bound}(T)](X^T)$; hence $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1) \equiv s(X_T) \land
B^m[\textsc{bound}(T)](X^T) \equiv \bot$.  Thus the second postcondition of
\textsc{\textoverline{learn}} holds.

>>>>

An abstraction of the game restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Figure~\ref{fig:agt} shows an
example abstract game tree for the controller (abstract game trees for the
environment are similar).  In the abstract game, the controller is required to
pick actions from the tree, starting from the root node.  After reaching a
leaf, it continues playing unrestricted.  The tree in Figure~\ref{fig:agt}
restricts the initial controller action to the set $\{a, d\}$.  After choosing
action $a$, the controller is required to play action $b$ in the next round.
It then reaches a leaf of the tree and continues playing unrestricted.  The
root of the tree is annotated by the initial state $s$ of the abstract game and
the bound $k$ on the number of rounds.  We denote $\textsc{nodes}(T)$ the set
of all nodes of a tree $T$, $\textsc{leaves}(T)$ the subset of leaf nodes, for
edge $e$ $\textsc{action}(e)$ is the action that labels the edge, and for node
$n$ $\textsc{height}(n)$ as the maximal distance from n to a leaf. For a node
$n$ of the tree, $\textsc{succ}(n)$ is the set of pairs $\langle e, n' \rangle$
where $n'$ is a child node of $n$ and $e$ is the edge connecting $n$ and $n'$.

@@

The tree is first extended to the maximum bound with edges that are labeled
with arbitrary opponent actions (Algorithm~\ref{alg:bounded}, line 14) as shown
in Figure~\ref{fig:extended}. Then copies of state and action variables are
introduced for each node in tree and opponent action variables for each edge
$e$ set to $\textsc{action}(e)$. The state and action variables of node $n$ are
connected to successor nodes $\textsc{succ}(n)$ by an encoding of the
transition relation and constrained to the winning condition of the player.

@@

\begin{algorithm}
    \caption{Tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{ $\lnot \Call{E}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{E}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{\Call{E}{$X_{T}$}}
        \Else
        \State \Return{ $\Call{E}{X_{T}} \lor$ \\
        $$\bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Require $s(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $s(X_T) \land B^M$
        \Function{learn}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min height 
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$s, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $s(X_T) \land B^m[$\Call{height}{$T$}$] \equiv \bot$
        \Function{\textoverline{learn}}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min height 
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{height}{$n$}}
                \State $B^m[i] \gets B^m[i] \setminus \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$s, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

@@

\begin{algorithm}[t]
    \caption{Amended tree formulas for Controller and Environment}
    \label{alg:unboundedTreeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{$B^M$}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{\Call{E}{$X_T$}}
        \Else
        \State \Return{ \Call{$B^m[\Call{height}{T}]$}{$X_{T}$} $\land$ \\
            $$\Call{E}{X_T} \lor \bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

Learning of states losing for the environment is similar (\textsc{\textoverline{learn}} in Algorithm~\ref{alg:learn}). 
The main difference is that environment-losing states are only losing for a particular
bound: given that there does not exist an environment strategy that
forces the game into an error state in $k$ rounds or less; there may still exist a longer
environment-winning strategy.  We therefore record learned environment-losing states
along with associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.  $B^m[k]$ 
are initialized to $\top$ for all $k$.  Whenever an environment-losing set $\mathcal{I}$
is discovered for a node $n$ with bound $\textsc{height}(n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is subtracted from 
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

@@

Formally, $\II$ has the property $\II \land F_2 \equiv \bot$.  Since $T_2$ is
of depth 1, this means that the environment cannot force the game into
$B^m[\textsc{height}(n)]$ playing against the counterexample moves in $T_2$.
Hence, $\II \cap Upre(B^m[\textsc{height}(n)]) = \emptyset$.  Furthermore, since
the may-invariant holds, $\II \cap Upre(i) = \emptyset$, for all $i <
\textsc{height}(n)$.  Hence, removing $\II$ from all $B^m[i], i\leq
\textsc{height}(n)$ in line~(15) preserves the may-invariant, thus satisfying the
first post-condition.

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ also satisfies $\II$.  Hence,
removing $\II$ from $B^m[\textsc{height}(n)]$ makes $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.

At the second last recursive call to \textsc{\textoverline{learn}}, tree $T_1$
is empty, $n$ is the root node, $\textsc{\textoverline{treeFormula}}(T_1)
\equiv B^m[\textsc{height}(T)](X^T)$; hence $s(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1) \equiv s(X_T) \land
B^m[\textsc{height}(T)](X^T) \equiv \bot$.  Thus the second postcondition of
\textsc{\textoverline{learn}} holds.
